---
title: "Pima Native Americans and Diabetes"
subtitle: "Technical Appendix"
author: "Amalia, Cassidy, Daniel, Maggie"
date: "12/21/2018"
output: pdf_document
---

```{r, include=F}
require(tidyverse)
require(caret)
require(pROC)
data <- read.delim("diabetes.txt", header = F, sep = ",") # assumes file is in the same directory
names(data) <- c("PRG", "PLASMA", "BP", "THICK", "INSULIN", "BODY", "PEDIGREE", "AGE", "RESPONSE")
set.seed(1234)
```

## Introduction

This is the technical appendix submitted for the Final Project portion of Statistics 320: Statistics Communications with Prof. Pamela Matheson. The assignment was to (A) predict the probability that individual females have diabetes, and (B) detect subsets of characteristics that are at higher risk of diabetes. This technical appendix details the code we used in our analysis and the creation of the data visualizations that comprised our presentation. 
- Presentation slides can be found [here](rpubs.com/danielnjoo/s320).
- The Shiny app can be found [here](danielnjoo.shinyapps.io/Pima_Eval_Model/)
- The 3 page report can be found [here]().
- The 1 page handout can be found [here]().

## Preliminary Analysis

To begin, we looked at the response variable to confirm how many cases of type 2 diabetes there were in this sample. Our findgings matched the handout: 268 out of 768 women in this sample were positive for type two diabetes. We then constructed two side by side histograms based on outside data in order to contextualize these findings. 

```{r}
pima<-data %>% clean_names()
pima$response<-as.factor(pima$response)


table(pima$response)
df1 <- data.frame("Prevalence" = c(34.2,40.8,9.2,8.3), "Gender" = c("Men", "Women", "Men", "Women"), "Group" = c("Pima","Pima","All","All"))

df1 %>% 
ggplot(aes(x=Group,y=Prevalence,fill=factor(Gender))) +
  geom_bar(stat="identity",position="dodge") +
  scale_fill_manual(name="Gender",values=c("#619CFF", "#F8766D")) +
  xlab("Group") + ylab("Percent with T2 Diabetes") + ggtitle("Type 2 Diabetes Rates") +
  theme(title = element_text(size=42), axis.title.x = element_text(size=32), axis.title.y = element_text(size=32), legend.text = element_text(size=32), axis.text.x = element_text(size=20), axis.text.y = element_text(size=20))  
```

This graph revealed that our sample was relatively consistent with outside estimates of diabetes prevalence among Pima Native American women. [^1] Additionally, we were able to visualize how much higher the rate of type two diabetes was among Pima Native American people relative to the general population of the US as a whole.  

Next, we investigated each variable separately by looking at their individual distributions for the women in our sample with diabetes (response=1) and those without (response=0).

```{r}

# PRG
pima %>% ggplot(aes(x=prg, group = response, color = response)) + geom_density() + ggtitle("PRG")

# plasma
pima %>% ggplot(aes(x=plasma, group = response, color = response)) + geom_density() + ggtitle("Plasma")

# BP
pima %>% ggplot(aes(x=bp, group = response, color = response)) + geom_density() + ggtitle("BP")

# thick
pima %>% ggplot(aes(x=thick, group = response, color = response)) + geom_density() + ggtitle("Thick")

# insulin
pima %>% ggplot(aes(x=insulin, group = response, color = response)) + geom_density() + ggtitle("Insulin")

# body
pima %>% ggplot(aes(x=body, group = response, color = response)) + geom_density() + ggtitle("BMI")

# pedigree
pima %>% ggplot(aes(x=pedigree, group = response, color = response)) + geom_density() + ggtitle("Pedigree")

# age
pima %>% ggplot(aes(x=age, group = response, color = response)) + geom_density() + ggtitle("Age")
```

We noticed differences in distributions among all variables, but some more than others. In the next section we use models to investigate these variables further.

## Model

We used `caret`'s `gbm()` (gradient boost model) function to initially create a classifier on all the variables available. Training options consisted of 5-fold cross-validation with a 0.75 train-test split. Given the function's random nature, a seed was set in order to ensure reproducibility.

```{r, include=F}
data$RESPONSE2 <- factor(ifelse(data$RESPONSE==1,"Y","N"), levels=c("N","Y"))
outcomeName <- 'RESPONSE2'
predictorsNames <- names(data)[1:8]
splitIndex <- createDataPartition(data[,outcomeName], p = .75, list = FALSE, times = 1)
trainDF <- data[ splitIndex,]
testDF  <- data[-splitIndex,]
objControl <- trainControl(method='cv', number=5, returnResamp='none', summaryFunction = twoClassSummary, classProbs = TRUE)
objModel <- train(trainDF[,predictorsNames], trainDF[,outcomeName], 
                  method='gbm', 
                  trControl=objControl,  
                  metric = "ROC",
                  preProc = c("center", "scale"))
```

A summary of the model object gives us the relative influences of the variables.

```{r}
summary(objModel)
```

We make predictions on the text set, plot an AUC curve, print the AUC, and the some overall metrics:

```{r}
predictions <- predict(object=objModel, testDF[,predictorsNames], type='prob')
auc <- roc(testDF$RESPONSE2, predictions[,2])
plot(auc); auc$auc
pred<-factor(ifelse(predictions$Y>0.50,"Y","N"),levels=c("N","Y"))
conf<-confusionMatrix(pred,testDF$RESPONSE2) 
conf$overall; conf$byClass
```

We find that the full variable model has an AUC of 85%, overall accuracy of 79%, sensitivity of 85%, and specificity of 69%. This full variable model is implemented in the Shiny app linked in the Introduction section.

### Two variable model

Recognizing that some of the variables such as Blood Plasma and Insulin would not be immediately available to epidemiologists on the ground surveying US Pima Native Americans, we felt the need to produce a reduced model as well. We ended up producing a 2 variable model consisting of BMI and Age, that still has a decent sensitivity (80%), but suffers from a low specificity (51%). Overall model accuracy was 70%.  

```{r, include=F}
predictorsNames2 <- predictorsNames[c(6,8)]
objModel2 <- train(trainDF[,predictorsNames2], trainDF[,outcomeName], 
                  method='gbm', 
                  trControl=objControl,  
                  metric = "ROC",
                  preProc = c("center", "scale"))
summary(objModel2)
predictions2 <- predict(object=objModel2, testDF[,predictorsNames], type='prob')
pred2<-factor(ifelse(predictions2$Y>0.50,"Y","N"),levels=c("N","Y"))
conf2<-confusionMatrix(pred2,testDF$RESPONSE2) 
```

```{r}
conf2$overall; conf2$byClass
```


Cassidy
```{r}
library(tidyverse)
pima <- data

pima <- pima %>% mutate(RESPONSE = ifelse(RESPONSE == 0, "No", "Yes"))

group.colors <- c(No = "cornflowerblue", Yes = "firebrick3")

pima %>% ggplot(aes(x=RESPONSE, y = PLASMA, fill = RESPONSE, group = RESPONSE)) + geom_boxplot() + ggtitle("Plasma Levels by Diabetes Status") + scale_x_discrete(name = "Tested positive for diabetes?") + scale_fill_manual(values = group.colors) + theme(legend.position="none", axis.text=element_text(size=12),
axis.title=element_text(size=14,face="bold"), plot.title = element_text(size = 18, hjust = 0.5)) + ylab("Plasma Glucose Concentration")

pima %>% ggplot(aes(x = RESPONSE, y = PEDIGREE, fill = RESPONSE, group = RESPONSE)) + geom_boxplot() + ggtitle("Pedigree by Diabetes Status") + scale_x_discrete(name = "Tested positive for diabetes?") + scale_fill_manual(values = group.colors) + theme(legend.position="none", axis.text=element_text(size=12),
axis.title=element_text(size=14,face="bold"), plot.title = element_text(size = 18, hjust = 0.5))

pima %>% ggplot(aes(x = RESPONSE, y = AGE, fill = RESPONSE, group = RESPONSE)) + geom_boxplot() + ggtitle("Diabetes Status by Age") + scale_x_discrete(name = "Tested positive for diabetes?") + scale_fill_manual(values = group.colors) + theme(legend.position="none", axis.text=element_text(size=12), axis.title=element_text(size=14,face="bold"), plot.title = element_text(size = 18, hjust = 0.5)) + ylab("Age (years)") + ylab("Pedigree")

pima %>% ggplot(aes(x = RESPONSE, y = BODY, fill = RESPONSE, group = RESPONSE)) + geom_boxplot() + ggtitle("BMI by Diabetes Status") + scale_x_discrete(name = "Tested positive for diabetes?") + scale_fill_manual(values = group.colors) + theme(legend.position="none") + ylab("BMI (Weight/Height)") + theme(axis.text=element_text(size=12),
axis.title=element_text(size=14,face="bold"), plot.title = element_text(size = 18, hjust = 0.5))

pima %>% 
  filter(BODY!=0) %>% 
  mutate(BODY=ifelse(BODY < 18.5, "underweight", 
         ifelse(BODY > 40, "morbid", ifelse(BODY > 30, "obese", ifelse(BODY > 25, "overweight", "normal"))))) %>%
  group_by(BODY, RESPONSE) %>% 
  summarize(n = n()) %>% 
  mutate(PERCENT = (n / sum(n))*100) %>% data.frame() -> temp


temp %>% 
  filter(RESPONSE=="Yes") %>% 
  ggplot(aes(BODY,  PERCENT)) + geom_bar(stat='identity', fill = "firebrick3") + 
  scale_x_discrete(limits=c("normal","overweight","obese","morbid"), labels=c("Normal","Overweight","Obese","Morbid")) + xlab("Weight Class (by BMI)") + 
  ylab("% of Positive Diabetes Test Results") + 
  theme(axis.text=element_text(size=15), axis.title=element_text(size=16,face="bold"), 
        plot.title = element_text(size = 18, hjust = 0.5)) + 
  ggtitle("% of Positive Test Results by Weight Class") + coord_cartesian(ylim=c(0,60))


pima %>% 
  mutate(AGE = ifelse(AGE < 30, "<30", 
         ifelse(AGE < 40, "30-49", ifelse(AGE < 50, "40-59", ifelse(AGE < 60, "60-69", "70+"))))) %>%
  group_by(AGE, RESPONSE) %>% 
  summarize(n = n()) %>% 
  mutate(PERCENT = (n / sum(n))*100) %>% data.frame() -> temp1


temp1 %>% 
  filter(RESPONSE=="Yes") %>% 
  ggplot(aes(AGE,  PERCENT)) + geom_bar(stat='identity', fill = "firebrick3") + 
  xlab("Age Group") + 
  ylab("% of Positive Diabetes Test Results") + 
  theme(axis.text=element_text(size=15), axis.title=element_text(size=16,face="bold"), 
        plot.title = element_text(size = 18, hjust = 0.5)) + 
  ggtitle("% of Positive Test Results by Age Group")

pima %>% 
  mutate(BP = ifelse(BP < 60, "Low", 
         ifelse(BP < 80, "Ideal", ifelse(BP < 90, "Pre-High", "High")))) %>%
  group_by(BP, RESPONSE) %>% 
  summarize(n = n()) %>% 
  mutate(PERCENT = (n / sum(n))*100) %>% data.frame() -> temp2

temp2 %>% 
  filter(RESPONSE=="Yes") %>% 
  ggplot(aes(BP,  PERCENT)) + geom_bar(stat='identity', fill = "firebrick3") + 
  scale_x_discrete(limits=c("Low","Ideal","Pre-High","High")) + xlab("Blood Pressure") + 
  ylab("% of Positive Diabetes Test Results") + 
  theme(axis.text=element_text(size=15), axis.title=element_text(size=16,face="bold"), 
        plot.title = element_text(size = 18, hjust = 0.5)) + 
  ggtitle("% of Positive Test Results by Blood Pressure") + coord_cartesian(ylim=c(0,60))

pima %>% 
  mutate(PLASMA = ifelse(PLASMA < 100, "Normal", 
         ifelse(PLASMA < 125, "Pre-Diabetic", "Diabetic"))) %>%
  group_by(PLASMA, RESPONSE) %>% 
  summarize(n = n()) %>% 
  mutate(PERCENT = (n / sum(n))*100) %>% data.frame() -> temp3

temp3 %>% 
  filter(RESPONSE=="Yes") %>% 
  ggplot(aes(PLASMA,  PERCENT)) + geom_bar(stat='identity', fill = "firebrick3") + 
  scale_x_discrete(limits=c("Normal","Pre-Diabetic","Diabetic")) + xlab("Blood Glucose Reading") + 
  ylab("% of Positive Diabetes Test Results") + 
  theme(axis.text=element_text(size=15), axis.title=element_text(size=16,face="bold"), 
        plot.title = element_text(size = 18, hjust = 0.5)) + 
  ggtitle("% of Positive Test Results by Blood Glucose Reading") + coord_cartesian(ylim=c(0,60))
```


## Data Visualization

In order to visualize the variables we deemed important, we split the sample into two groups: women who were positive for type 2 diabetes and women who were negative. Then, we constructed overlapping histograms for each of the four variables we were interested in so that we could see how the distributions for each variable differed between groups. This is similar to what we did in the preliminary analysis section; the only difference (in addition to narrowing down the number of variables we're presenting) is we have made the graphs more visually appealing and easy to understand for someone unfamiliar with the dataset. These are the visuals we used in our presentation.

```{r}
#install.packages("rgr")
require(rgr)
require(lattice)
require(janitor)
require(ggpubr)

d1<-data %>% clean_names()

d2<-filter(d1, response==1)%>%mutate(Diabetes=factor(ifelse(response==1, "Positive", NA)))#positive
d3<-filter(d1, response==0)%>%mutate(Diabetes=factor(ifelse(response==0, "Negative", NA)))#negative

p0 <- ggplot(data.frame()) + geom_blank() + theme(panel.background=element_blank())

p1 <- ggplot(d2,aes(x=plasma)) + 
  geom_histogram(data=d2,aes(x=plasma, fill=Diabetes), alpha = 0.7) +
  geom_histogram(data=d3,aes(x=plasma, fill=Diabetes), alpha = 0.7) + theme(plot.title=element_text(hjust=0.5)) + labs(x="Plasma Glucose Concentration", y="Number of People", title="Plasma") +  theme(title = element_text(size=16), legend.position = "none")

p2 <- ggplot(d2,aes(x=age)) + 
  geom_histogram(data=d2,aes(x=age, fill=Diabetes), alpha = 0.7)+
  geom_histogram(data=d3,aes(x=age, fill=Diabetes), alpha = 0.7) + theme(plot.title=element_text(hjust=0.5)) + labs(x="Age (In Years)", y="Number of People", title="Age")+  theme(title = element_text(size=16), legend.position="none")

p3 <- ggplot(d2,aes(x=pedigree)) + 
  geom_histogram(data=d2,aes(x=pedigree, fill=Diabetes), alpha = 0.7)+
  geom_histogram(data=d3,aes(x=pedigree, fill=Diabetes), alpha = 0.7)+ theme(plot.title=element_text(hjust=0.5)) + labs(x="Pedigree Function Value", y="Number of People", title="Pedigree")+  theme(title = element_text(size=16), legend.position = "none")

p4 <- ggplot(d2,aes(x=body)) + 
  geom_histogram(data=d2,aes(x=body, fill=Diabetes), alpha = 0.7)+
  geom_histogram(data=d3,aes(x=body, fill=Diabetes), alpha = 0.7)+ theme(plot.title=element_text(hjust=0.5)) + labs(x="BMI", y="Number of People", title="Body Mass Index")+  theme(title = element_text(size=16), legend.position = "bottom")

ggarrange(p1, p2, p3, p4, ncol=2, nrow=2, common.legend = TRUE, legend="bottom")
```

As you can see, women in our sample who tested positive for type 2 diabetes trended higher for each variable in question (plasma, age, pedigree, BMI). This suggests that a high plasma glucose concentration, greater age, a high pedigree value, and a high BMI may all be risk factors for Type 2 diabetes.

[^1]: "Effects of Traditional and Western Environments on Prevalence of Type 2 Diabetes in Pima Indians in Mexico and the U.S." 2006. Leslie Schulz, Peter Bennett, Eric Ravussin, Judith Kidd, Kenneth Kidd, Julian Esparza and Mauro E. Valencia.
http://care.diabetesjournals.org/content/29/8/1866